---
title: "DATA 589 Project - Single covariate modelling - Distance to Water"
author: "Amali Jayatileke"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document:
    toc: true
    theme: cerulean
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Single Covariate Analysis for 'Distance to Water'

```{r, include=FALSE}
#load packages
library(sp)
library(sf)
library(spatstat)
library(rgbif)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(raster)
library(mgcv)
```

### Load bears dataset

```{r}
# # get occurrences from GBIF
#bc_occ <- occ_search(
#   scientificName = "Ursus americanus",
#   country = "CA",
#   stateProvince = "British Columbia",
#   limit = 5000
# )
 
 # omit records with missing coordinates
#occ_data <- bc_occ$data[!is.na(bc_occ$data$decimalLongitude) & !is.na(bc_occ$data$decimalLatitude), ]
 
# save occurrences
#save(occ_data, file = "bc_bear_occurrences2.Rda")
```

```{r}
# load and view data
load("bc_bear_occurrences.Rda")
str(occ_data)
```

### Load covariates

```{r}
load("BC_Covariates.Rda")
summary(DATA)
```

### Create ppp object

```{r}
# extract location columns
bears_loc <- occ_data[, c("decimalLongitude", "decimalLatitude", "month", "year")]
bears_loc_filtered <- subset(bears_loc, year %in% c(2020, 2021, 2022, 2023, 2024))
#bears_loc_filtered <- subset(bears_loc, year %in% c(2024))

# create sf object
bears_sf <- st_as_sf(
  bears_loc_filtered,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326  # WGS84 (longitude/latitude)
)

# BC Albers projection
bears_sf_proj <- st_transform(bears_sf, crs = 3005)

# extract x, y coordinates
coords <- st_coordinates(bears_sf_proj)

# extract BC window
window_sf <- st_as_sf(DATA$Window) # convert SpatialPolygons to Simple Features (sf)
window_proj <- st_transform(window_sf, crs = 3005) # ensure same CRS
window <- as.owin(window_proj) # convert to owin using sf object

# To avoid plotting points outside window (illegal points)
# Intersect points with window
#bears_sf_proj_filtered <- bears_sf_proj[st_within(bears_sf_proj, window_proj, sparse = FALSE), ]
bears_sf_proj <- subset(bears_sf_proj, st_within(bears_sf_proj, window_proj, sparse = FALSE))

# Re-extract coordinates
coords_filtered <- st_coordinates(bears_sf_proj)

# Create new ppp object
bears_ppp <- ppp(
 x = coords_filtered[, 1],
 y = coords_filtered[, 2],
 window = window
)

#plot(bears_ppp, pch = 21, main = "Black bear occurrences in BC, 2020 - 2024")

```

### Create ppp objects for seasons

```{r}
# Define the seasons and corresponding colors
seasons <- list(
  winter = c(12, 1, 2),
  spring = c(3, 4, 5),
  summer = c(6, 7, 8),
  autumn = c(9, 10, 11)
)

# Create an empty list to store the ppp objects for each season
ppp_list <- list()

# Create an empty list to store the filtered data for each season
bears_sf_list <- list()

# Loop over seasons
for (i in 1:length(seasons)) {
  
  # Filter for each season
  season_name <- names(seasons)[i]
  season_months <- seasons[[i]]
  
  # Filter bears_loc_filtered by the season's months
  bears_loc_season <- bears_loc_filtered[bears_loc_filtered$month %in% season_months, ]
  
  # Print the number of observations for this season
  cat(season_name, ": ", nrow(bears_loc_season), " observations\n", sep="")
  
  # Create sf object for the filtered season
  bears_sf_season <- st_as_sf(
    bears_loc_season,
    coords = c("decimalLongitude", "decimalLatitude"),
    crs = 4326  # WGS84 (longitude/latitude)
  )
  
  # Transform to BC Albers projection
  bears_sf_season_proj <- st_transform(bears_sf_season, crs = 3005)
  
  # Store sf object in the list
  bears_sf_list[[season_name]] <- bears_sf_season_proj
  
  # Extract x, y coordinates
  coords <- st_coordinates(bears_sf_season_proj)
  
  # Create ppp object for each season
  suppressWarnings(
    bears_ppp_season <- ppp(
      x = coords[, 1],
      y = coords[, 2],
      window = window
    )
  )
  
  # Store ppp object in the list
  ppp_list[[season_name]] <- bears_ppp_season
}
```


### EDA

```{r}
elev <- na.omit(DATA$Elevation)
cover <- na.omit(DATA$Forest)
dist_water <- na.omit(DATA$Dist_Water)
hfi <- na.omit(DATA$HFI)

```

```{r}
is.im(dist_water)          # should return TRUE
is.ppp(bears_ppp)          # should return TRUE

```

#### Intensity plots

```{r, cache = TRUE}
# Overall
# intensity as a function of distance from water
rho <- rhohat(bears_ppp, dist_water) 
plot(rho, xlim=c(0, max(dist_water)), main = "Estimated rho vs Distance to Water", xlab = "Distance to water")

```

**Interpretation:**

-   Higher Intensity Close to Water

    -   At **short distances (0–2000 meters)**, the black line is **above** the red dashed line. This suggests that **bear occurrences are more likely near water** than average.

-   **Decreasing Intensity With Distance**

    -   As distance to water increases (beyond \~5000 meters), the relative intensity generally **declines**, often dipping below the mean (red dashed line). This means **bears are less likely to be found farther from water.**

-   **Multiple Peaks at Mid-Range Distances**

    -   There are small **bumps at \~3000, \~8000, and \~13000 meters**. These could indicate:

    -   Presence of **habitat patches** or **water-proximate features** that bears favor.

    -   **Random placement** of some bear points that just happen to cluster around a certain distance (e.g., \~10,000m from water) purely by chance.

    -   **Noise** or **measurement errors**

    -   Small sample size, bumps after 5000 m have larger confidence bands indicating high uncertainty due to smaller number of data point.

    -   Possibly transitional areas like wetlands or secondary water sources.

-   **Wide Confidence Band Beyond \~15000m**

    -   The grey band becomes **very wide** beyond 15,000 meters, and the estimate drops near zero.

    -   **Very few data points** (bear locations or quadrature points) are this far from water.

-   **Conclusion** Bear occurrences are strongly concentrated near water. There is a clear negative relationship between bear intensity and distance to water: the farther from water, the less likely a bear is to be found. This supports the ecological hypothesis that proximity to water is a key driver of black bear spatial distribution in BC.

```{r, cache = TRUE, fig.width = 10, fig.height = 5}
par(mfrow = c(2,2), mar = c(4, 4, 3, 2))
for (i in 1:length(ppp_list)) {
  season_name <- names(ppp_list)[i]
  rho <- rhohat(ppp_list[[season_name]], dist_water) 
  plot(rho, legend = FALSE, xlim = c(0, max(dist_water)), main = paste(season_name))
  legend("topright", legend = c(expression(hat(rho)(dist_water)), expression(bar(rho))), 
         col = c("black", "red"), lty = c(1, 2), bty = "n")
}

```

**Interpretation**

1.  Winter

-   Moderate positive intensity at 3,000–8,000m, peaking slightly around 7,000–8,000m.

-   Confidence bands widen after \~8,000m, suggesting fewer observations.

-   Little to no bear activity at farther distances (\>15,000m).

**Interpretation:** Bears may be located in moderately water-proximate areas during winter, possibly due to denning or reduced mobility.

2.  Spring

-   Sharp initial rise, fluctuating but generally high intensity between 0–10,000m.

-   Peaks between 2,000–5,000m and 8,000–10,000m.

-   Wider confidence intervals suggest less certainty in the far range.

**Interpretation:**Spring shows stronger proximity preference to water, likely due to post-hibernation activity, foraging, or mating behavior.

3.  Summer

-   Very strong intensity peaks near 7,000m and 12,000–14,000m.

-   Complex structure—multiple peaks and valleys.

-   Wider confidence intervals indicate heterogeneity in behavior or data sparsity.

**Interpretation:** Summer suggests localized hotspots of bear activity, possibly reflecting feeding grounds or specific habitat zones near water.

4.  Autumn

-   Overall high intensity near 2,000–6,000m, then multiple sharp peaks at 9,000–14,000m.

-   Some intensity estimates exceed the mean line, indicating strong preference in those zones.

**Interpretation:** Bears may be preparing for hibernation and seeking resources—like streams or food patches near water—resulting in varied but elevated usage across mid-range distances.

**Key Takeaways:**

-   Bears show non-linear, seasonal-specific relationships with distance to water.

-   There's no consistent monotonic trend; instead, multiple peaks suggest preferred zones or habitat patchiness.

-   Confidence bands widen beyond \~10–12 km, so interpretation is less reliable at far distances.

```{r, cache = TRUE}
# Seasonal 
# intensity as a function of distance from water
#par(mfrow=c(2,2), mar=c(2,2,2,2))
#for (i in 1:length(ppp_list)) {
#  season_name <- names(ppp_list)[i]
#  rho <- rhohat(ppp_list[[season_name]], dist_water) 
#  plot(rho, xlim=c(0, max(dist_water)), main = paste(season_name))
#}
```

```{r}
# Distance to water for bears locations

# im object to RasterLayer
im2raster <- function(im_obj) {
  # iamge to matrix
  m <- as.matrix(im_obj)
  # create raster object
  r <- raster(m)
  # range and extension
  ext <- c(im_obj$xrange[1] - im_obj$xstep/2,
           im_obj$xrange[2] + im_obj$xstep/2,
           im_obj$yrange[1] - im_obj$ystep/2,
           im_obj$yrange[2] + im_obj$ystep/2)
  extent(r) <- ext
  return(r)
}

# dist_water to RasterLayer
distw_raster <- im2raster(dist_water)

# dist_water in bear locations
bears_sf_proj$distw_value <- raster::extract(distw_raster, bears_sf_proj) #For each bear point, extract the distance to water value from the raster.

summary(bears_sf_proj$distw_value)
```

```{r}
hist(bears_sf_proj$distw_value, 
     main = "Distance to Water at bear locations", 
     xlab = "Distance to Water",
     col = "lightblue", border = "grey")

```

```{r}
distw_values <- getValues(distw_raster)

hist(distw_values, 
     main = "Global distribution of Distance to Water in BC", 
     xlab = "Distance to Water",
     col = "lightblue3", border = "grey")
```

### Distance to Water effects

```{r}

# Modelling without raster object
# model - Linear
mod_distw_clean <- ppm(bears_ppp, ~ dist_water, covariates = list(dist_water))
mod_distw_clean
```

```{r}

#Plot the model predictions
plot(mod_distw_clean,
     se = FALSE,
     superimpose = FALSE,
     box = F,
     main = "Model predictions - Linear effects")

#Overlay the bear occurences
plot(bears_ppp,
     pch = 16,
     cex = 0.7,
     cols = "white",
     add = TRUE)
plot(bears_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)

```

```{r}

# Modelling without raster object
# model - Quadratic
mod_distw_clean_quad <- ppm(bears_ppp, ~ dist_water + I(dist_water^2), covariates = list(dist_water))
mod_distw_clean_quad
```

```{r}

#Plot the model predictions
plot(mod_distw_clean_quad,
     se = FALSE,
     superimpose = FALSE,
     box = F,
     main = "Model predictions - Quadratic effects")

#Overlay the bear occurences
plot(bears_ppp,
     pch = 16,
     cex = 0.7,
     cols = "white",
     add = TRUE)
plot(bears_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)

```

```{r}
library(mgcv)  # for spline functionality

# Fit a non-linear model using a smooth function of dist_water
mod_distw_smooth <- ppm(bears_ppp,
                        ~ s(dist_water),
                        covariates = list(dist_water = dist_water),
                        use.gam = TRUE)
mod_distw_smooth
```

```{r}

#Plot the model predictions
plot(mod_distw_smooth,
     se = FALSE,
     superimpose = FALSE,
     box = F,
     main = "Model predictions - Smooth effects(GAM)")

#Overlay the bear occurences
plot(bears_ppp,
     pch = 16,
     cex = 0.7,
     cols = "white",
     add = TRUE)
plot(bears_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)

```

```{r}
# model - Null
mod_distw_clean_null <- ppm(bears_ppp, ~ 1)
mod_distw_clean_null


```

#### Model evaluation

```{r}
# Compare models using AIC

# Null model
cat("AIC value for the null model:",AIC(mod_distw_clean_null))

# Linear model
cat("\nAIC value for the linear effects model:",AIC(mod_distw_clean))

# Quadratic model
cat("\nAIC value for the quadratic effects model:",AIC(mod_distw_clean_quad))

# GAM-style Smooth model
cat("\nAIC value for the GAM Smooth effects model:",AIC(mod_distw_smooth),"\n")

AIC(mod_distw_clean) - AIC(mod_distw_clean_quad)
```

**Interpretation**

-   Linear model has lowest AIC among the 4 models fitted. Therefore, fitted model with **linear effect of covariate** (`mod_distw_clean`) is the better model

-   However note that the quadratic term was not significant.

```{r}
options(scipen = 0)

# Likelihood Ratio Test 

# Null model vs Linear effects
anova(mod_distw_clean_null, mod_distw_clean,  test = "LRT")

# Null model vs Quadratic effects
anova(mod_distw_clean_null, mod_distw_clean_quad,  test = "LRT")

# Likelihood Ratio Test 
anova(mod_distw_clean, mod_distw_clean_quad,  test = "LRT")
```

**Interpretation**

-   This performs a likelihood ratio test under the null hypothesis that the simpler model is sufficient.
-   The p-value (\< 2.2e-16) is very small (\< 0.05) when comparing the linear and quadratic effects with the null model. However the p-value is 0.4022 (\>0.05) when comparing the linear vs quadratic effects models, therefore, the **linear effects model is best fit**.

```{r}
# equation for the selected model
options(scipen = 999) # remove scientific notation
mod_distw_clean$coef

```

$$
\lambda_{bears}(u) = e^{-19.36334426884 -0.00005085943.distwater(u)}
$$

```{r}
# Test whether the observed data deviate significantly from the model predictions.
# Run the quadrat test
quadrat.test(mod_distw_clean, nx = 2, ny = 2)
```

**Interpretation** - The small p value( \< 2.2e-16) indicates that observed data deviate significantly from the model predictions.

```{r}
#Calculate the residuals
res <- residuals(mod_distw_clean)

#Visualise
plot(res,
     main = "Model residuals",
     cols = "transparent"
     )
```

```{r}
#Calculate the residuals
res <- residuals(mod_distw_clean)

#Visualise
plot(res,
     main = "Model residuals"
     )
```

```{r}
mod_res_clean <- residuals(mod_distw_clean, type = "pearson")

mod_res_clean$v[!is.finite(mod_res_clean$v)] <- NA

# Residuals
plot(mod_res_clean, main = "Model Residual Map (Pearson)", na.col = "transparent")
```


**Interpretation**

-   Since the residual plot shows no obvious patterns in the residuals, we can conclude the model is accurately capturing the trends in the dataset.

```{r}
#Calculate the partial residuals as a function of distance to water
par_res_distw <- parres(mod_distw_clean, "dist_water")

plot(par_res_distw,
     legend = FALSE,
     lwd = 2,
     main = "Partial residuals Distance to water - All seasons",
     ylab = 'Partial Residuals',
     xlab = "Distance to water")
```

### Seasonal Distance to Water Effects

```{r}
options(scipen = 999) # remove scientific notation

# model for each season (individually)
models_seasonal <- list()

for (season in names(ppp_list)) {
  models_seasonal[[season]] <- ppm(ppp_list[[season]], ~ dist_water, covariates = list(dist_water))
  cat("Model for", season, ":\n")
  print(models_seasonal[[season]])
}
```

**Interpretation:**

1.  **Winter**

-   Intercept: significant

-   Slope (dist_water): (p \> 0.05) not significant

-   **Interpretation:**

    -   Bear occurrence does not significantly vary with distance to water. The flat slope suggests distance to water is not an important factor in winter — possibly because bears are less mobile due to hibernation.

2.  **Spring**

-   Intercept: significant

-   Slope (dist_water): -4.26e-05, p ≈ 0.05 (\*)

-   **Interpretation:**

    -   There's weak evidence that bear occurrence decreases with increasing distance from water. Bears may start to rely on water sources as they emerge from hibernation and become active.

3.  **Summer**

-   Intercept: significant

-   Slope (dist_water): p \< 0.05 (\*)

-   **Interpretation:**

    -   Bear occurrence significantly decreases with distance from water.This is ecologically expected, as summer means more activity, foraging, and need for hydration.

4.  **Autumn**

-   Intercept: -20.72 (significant)

-   Slope (dist_water): -1.13e-04, p \< 0.001 (\*\*\*)

-   **Interpretation:**

    -   Bears are strongly associated with areas closer to water.This is the strongest seasonal effect — likely due to pre-hibernation foraging and increased movement and calorie intake near water sources.

#### Model Evaluation

```{r}
# Calculate partial residuals for each season's model
partial_residuals_seasonal <- list()

for (season in names(models_seasonal)) {
  partial_residuals_seasonal[[season]] <- parres(models_seasonal[[season]], covariate = "dist_water")
}
```

```{r, fig.width=10, fig.height=5}
# Plot partial residuals in a 2x2 grid
par(mfrow = c(2, 2), mar = c(4, 4, 2, 2))  # Larger margins for axes and labels

for (i in seq_along(partial_residuals_seasonal)) {
  season_name <- names(partial_residuals_seasonal)[i]
  plot(
    partial_residuals_seasonal[[i]],
    main = paste("Partial Residuals Distance to Water\n", season_name),
    xlab = "Distance to Water",
    ylab = 'Partial Residuals',
    legend = F
  )
}
```

