---
title: "Ursus Americanus - Forest Cover Analysis"
author: "Jane Shen"
date: "2025-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# load packages
library(sp)
library(sf)
library(spatstat)
library(rgbif)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(raster)
library(mgcv)
```

```{r}
# load data
load("bc_bear_occurrences.Rda")
load("BC_Covariates.Rda")
```

### Estimate intensity of bear occurrences based on forest cover

```{r}
# intensity as a function of forest cover overall
cover <- DATA$Forest
rho <- rhohat(bears_ppp, cover) 
plot(rho, 
     xlim=c(0, max(cover)), 
     main="Estimated Rho vs. Forest Cover")
```  

```{r}
# intensity as a function of forest cover broken up by seasons
par(mfrow=c(2,2), mar=c(2,2,2,2))

for (i in 1:length(ppp_list)) {
  season_name <- names(ppp_list)[i]
  rho <- rhohat(ppp_list[[season_name]], cover)
  plot(rho, 
       xlim=c(0, max(cover)), 
       main = paste(season_name))
}
```

```{r}
# convert im to raster
im2raster <- function(im_obj) {
  m <- as.matrix(im_obj) # image to matrix
  r <- raster(m) # create raster object
  
  # range and extension
  ext <- c(im_obj$xrange[1] - im_obj$xstep/2,
           im_obj$xrange[2] + im_obj$xstep/2,
           im_obj$yrange[1] - im_obj$ystep/2,
           im_obj$yrange[2] + im_obj$ystep/2)
  extent(r) <- ext
  return(r)
}

# extract forest cover values at occurrence locations
cover_raster <- im2raster(cover)
bears_sf_proj$cover_value <- extract(cover_raster, bears_sf_proj)
```

#### KDE to estimate the concentration of bear occurrences based on forest cover values

```{r}
# KDE for Bear Locations
kde_b <- density(bears_sf_proj$cover_value, na.rm = TRUE, bw = "SJ-dpi")
summary(bears_sf_proj$cover_value)

plot(kde_b,
     main = "Kernel Density Estimate of Forest Cover at Bear Locations",
     xlab = "Forest Cover (%)",
     ylab = "Density",
     col = "#2C6B2F", 
     lwd = 2)  
```

These results suggest that bears are found in a mix of forested and non-forested areas, but on average, they’re in areas with moderate to high forest cover.

### Model Forest Cover

```{r}
# NA with mean forest cover
mean_cover <- mean(as.vector(as.matrix(cover)), na.rm = TRUE)
cover_clean <- eval.im(ifelse(is.na(cover), mean_cover, cover))

# polynomial model
cover_model <- ppm(bears_ppp ~ cover + I(cover^2), covariates = list(cover = cover_clean))
print(cover_model)

# predicts and plots intensity
pred_intensity_cover <- predict(cover_model)
plot(pred_intensity_cover, 
     main = "Predicted Intensity Based on Forest Cover")
points(bears_ppp, 
       pch = 20, 
       col = "red")
```
The positive coefficient on the linear term indicates that bear occurrence intensity initially increases with forest cover. However, the negative coefficient on the quadratic term means that this relationship has a peak, after which further increases in forest cover are associated with a decrease in expected intensity. This suggests that bears are most likely to be found in areas with moderate forest cover, and less likely to occur in areas with either very low or very high forest density. 

This pattern may reflect ecological preferences, where bears favor mixed habitats over open or densely forested areas. It’s also possible that human detection rates are lower in areas of dense vegetation, where visibility is reduced.

#### Compute and plot residuals

```{r}
cover_residuals <- residuals(cover_model, type = "pearson")

# remove any non-finite residuals
cover_residuals$v[!is.finite(cover_residuals$v)] <- NA

# plot forest cover residuals
plot(cover_residuals, 
     main = "Residuals - Forest Cover Model", 
     na.col = "transparent")
```

#### Partial Residuals

```{r}
# Computes partial residuals for forest cover
par_res_cover_all <- parres(cover_model, "cover")

# Plots partial residuals
plot(par_res_cover_all,
     legend = FALSE,
     lwd = 2,
     main = "Partial Residuals Forest Cover - All Seasons",
     xlab = "Forest Cover (%)",
     ylab = "Partial Residuals")
```

### Seasonal Models  

```{r}
models_cover_seasonal <- list()

# model for each season (individually)
for (season in names(ppp_list)) {
  models_cover_seasonal[[season]] <- ppm(ppp_list[[season]], ~ cover + I(cover^2), covariates = list(cover = cover_clean))
  # cat("Model for", season, ":\n")
  # print(models_seasonal[[season]])
  plot(predict(models_cover_seasonal[[season]]), main = paste("Intensity ~ Forest Cover -", season))
}
```  

**Winter: Winter: Intercept: -23.78; Forest Coefficient: 0.0582; Quadratic Coefficient: -0.000672.**

Bear occurrences increase with forest cover up to a point, then decline. The smaller linear coefficient suggests a slower rise in intensity compared to other seasons. The model did not converge, so caution is needed in interpreting the results.

**Spring: Intercept: -21.96; Forest Coefficient: 0.0657; Quadratic Coefficient: -0.000632.**

The relationship is similar to winter but slightly stronger, meaning a more noticeable increase and peak in moderate forest areas.

**Summer: Intercept: -21.05; Forest Coefficient: 0.0621; Quadratic Coefficient: -0.000602.**

Summer shows a comparable trend with a slightly weaker curvature, suggesting bears may tolerate higher forest density before intensity starts to drop.

**Autumn: Intercept: -21.70; Forest Coefficient: 0.0678; Quadratic Coefficient: -0.000732.**

Autumn has the sharpest curve (greatest linear and quadratic coefficient magnitudes), indicating a more distinct peak and a stronger decline in high-density forest areas.

Overall, all seasons display a consistent non-linear relationship where bear occurrence intensity increases with forest cover up to a moderate level, then declines. While the shape of the curve remains similar, the strength and location of the peak vary slightly by season, indicating subtle shifts in habitat preference throughout the year.

#### Partial Residuals per Season

```{r}
# 2x2 plot layout
par(mfrow = c(2, 2))

# Generates and plots partial residuals per season
for (season in names(models_cover_seasonal)) {
  model <- models_cover_seasonal[[season]]
  parres_season <- parres(model, "cover")
  plot(parres_season,
       legend = FALSE,
       lwd = 2,
       main = paste("Partial Residuals - Forest Cover", season),
       xlab = "Forest Cover (%)",
       ylab = "Partial Residuals")
}
```