---
title: "DATA 589 Project - Mulit-covariate modelling"
author: "Amali Jayatileke"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
  pdf_document:
    latex_engine: xelatex
---

# Multi-Covariate Analysis


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#load packages
library(sp)
library(sf)
library(spatstat)
library(rgbif)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(raster)
library(mgcv)
```

### Load bears dataset

```{r}
# # get occurrences from GBIF
#bc_occ <- occ_search(
#   scientificName = "Ursus americanus",
#   country = "CA",
#   stateProvince = "British Columbia",
#   limit = 5000
# )
 
 # omit records with missing coordinates
#occ_data <- bc_occ$data[!is.na(bc_occ$data$decimalLongitude) & !is.na(bc_occ$data$decimalLatitude), ]
 
# save occurrences
#save(occ_data, file = "bc_bear_occurrences2.Rda")
```

```{r}
# load and view data
load("bc_bear_occurrences.Rda")
str(occ_data)
```

### Load covariates

```{r}
load("BC_Covariates.Rda")
summary(DATA)
```

### Create ppp object

```{r}
# extract location columns
bears_loc <- occ_data[, c("decimalLongitude", "decimalLatitude", "month", "year")]
bears_loc_filtered <- subset(bears_loc, year %in% c(2020, 2021, 2022, 2023, 2024))
#bears_loc_filtered <- subset(bears_loc, year %in% c(2024))

# create sf object
bears_sf <- st_as_sf(
  bears_loc_filtered,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326  # WGS84 (longitude/latitude)
)

# BC Albers projection
bears_sf_proj <- st_transform(bears_sf, crs = 3005)

# extract x, y coordinates
coords <- st_coordinates(bears_sf_proj)

# extract BC window
window_sf <- st_as_sf(DATA$Window) # convert SpatialPolygons to Simple Features (sf)
window_proj <- st_transform(window_sf, crs = 3005) # ensure same CRS
window <- as.owin(window_proj) # convert to owin using sf object

# To avoid plotting points outside window (illegal points)
# Intersect points with window
#bears_sf_proj_filtered <- bears_sf_proj[st_within(bears_sf_proj, window_proj, sparse = FALSE), ]
bears_sf_proj <- subset(bears_sf_proj, st_within(bears_sf_proj, window_proj, sparse = FALSE))

# Re-extract coordinates
coords_filtered <- st_coordinates(bears_sf_proj)

# Create new ppp object
bears_ppp <- ppp(
 x = coords_filtered[, 1],
 y = coords_filtered[, 2],
 window = window
)

#plot(bears_ppp, pch = 21, main = "Black bear occurrences in BC, 2020 - 2024")

```

### Create ppp objects for seasons

```{r}
# Define the seasons and corresponding colors
seasons <- list(
  winter = c(12, 1, 2),
  spring = c(3, 4, 5),
  summer = c(6, 7, 8),
  autumn = c(9, 10, 11)
)

# Create an empty list to store the ppp objects for each season
ppp_list <- list()

# Create an empty list to store the filtered data for each season
bears_sf_list <- list()

# Loop over seasons
for (i in 1:length(seasons)) {
  
  # Filter for each season
  season_name <- names(seasons)[i]
  season_months <- seasons[[i]]
  
  # Filter bears_loc_filtered by the season's months
  bears_loc_season <- bears_loc_filtered[bears_loc_filtered$month %in% season_months, ]
  
  # Print the number of observations for this season
  cat(season_name, ": ", nrow(bears_loc_season), " observations\n", sep="")
  
  # Create sf object for the filtered season
  bears_sf_season <- st_as_sf(
    bears_loc_season,
    coords = c("decimalLongitude", "decimalLatitude"),
    crs = 4326  # WGS84 (longitude/latitude)
  )
  
  # Transform to BC Albers projection
  bears_sf_season_proj <- st_transform(bears_sf_season, crs = 3005)
  
  # Store sf object in the list
  bears_sf_list[[season_name]] <- bears_sf_season_proj
  
  # Extract x, y coordinates
  coords <- st_coordinates(bears_sf_season_proj)
  
  # Create ppp object for each season
  suppressWarnings(
    bears_ppp_season <- ppp(
      x = coords[, 1],
      y = coords[, 2],
      window = window
    )
  )
  
  # Store ppp object in the list
  ppp_list[[season_name]] <- bears_ppp_season
}
```


### EDA

```{r}
elev <- na.omit(DATA$Elevation)
cover <- na.omit(DATA$Forest)
dist_water <- na.omit(DATA$Dist_Water)
hfi <- na.omit(DATA$HFI)

```

```{r}
is.im(dist_water)          # should return TRUE
is.ppp(bears_ppp)          # should return TRUE

```

### Check Collinearity among covariates

```{r}
# Distance to water for bears locations

# im object to RasterLayer
im2raster <- function(im_obj) {
  # iamge to matrix
  m <- as.matrix(im_obj)
  # create raster object
  r <- raster(m)
  # range and extension
  ext <- c(im_obj$xrange[1] - im_obj$xstep/2,
           im_obj$xrange[2] + im_obj$xstep/2,
           im_obj$yrange[1] - im_obj$ystep/2,
           im_obj$yrange[2] + im_obj$ystep/2)
  extent(r) <- ext
  return(r)
}

# dist_water to RasterLayer
distw_raster <- im2raster(dist_water)

# dist_water in bear locations
bears_sf_proj$distw_value <- raster::extract(distw_raster, bears_sf_proj) #For each bear point, extract the distance to water value from the raster.

summary(bears_sf_proj$distw_value)
```

```{r}
elev_raster <- im2raster(elev)
cover_raster <- im2raster(cover)
hfi_raster <- im2raster(hfi)

crs(elev_raster) <- CRS("+init=epsg:3005") 
crs(cover_raster) <- CRS("+init=epsg:3005") 
crs(hfi_raster) <- CRS("+init=epsg:3005") 

bears_sf_proj$elev_value <- raster::extract(elev_raster, bears_sf_proj)
bears_sf_proj$cover_value <- raster::extract(cover_raster, bears_sf_proj)
bears_sf_proj$hfi_value <- raster::extract(hfi_raster, bears_sf_proj)


summary(bears_sf_proj$elev_value)
summary(bears_sf_proj$cover_value)
summary(bears_sf_proj$hfi_value)
```

```{r}
#Check for collinearity
cov_df <- data.frame(
  distw = bears_sf_proj$distw_value,
  elev = bears_sf_proj$elev_value,
  cover = bears_sf_proj$cover_value,
  hfi = bears_sf_proj$hfi_value
)
# Remove rows with NAs
cov_df_clean <- na.omit(cov_df)

# Check Correlation Matrix
cor_matrix <- cor(cov_df_clean)
print(round(cor_matrix, 2))

```

### Multi-covariate effects

```{r}
elev <- na.omit(DATA$Elevation)
cover <- na.omit(DATA$Forest)
dist_water <- na.omit(DATA$Dist_Water)
hfi <- na.omit(DATA$HFI)

# Problem:  Values of the covariate ‘hfi’ were NA or undefined at 0.32% (32 out of 9920) of the quadrature points, 
# when fitting spatial point process models with ppm() in spatstat, missing covariate values at quadrature points will cause errors. Hence convert to raster object

```

```{r}

# hfi to RasterLayer
hfi_raster <- im2raster(hfi)

hfi_matrix <- as.matrix(hfi)
media_hfi <- mean(as.vector(hfi_matrix), na.rm = TRUE)
cat("Mean HFI (without NA):", media_hfi, "\n")

# Impute NA values in HFI using the mean
hfi_clean <- eval.im( ifelse(is.na(hfi), media_hfi, hfi) )

summary(as.vector(as.matrix(hfi_clean)))


# elevation to RasterLayer
elev_raster <- im2raster(elev)

elev_matrix <- as.matrix(elev)
media_elev <- mean(as.vector(elev_matrix), na.rm = TRUE)
cat("Mean Elevation (without NA):", media_elev, "\n")

# Impute NA values in Elevation using the mean
elev_clean <- eval.im( ifelse(is.na(elev), media_elev, elev) )

summary(as.vector(as.matrix(elev_clean)))


```

```{r}
# model - Null
mod_distw_clean_null <- ppm(bears_ppp, ~ 1)
mod_distw_clean_null


```

```{r}

# Model 1:  Linear effects of all 4 covariates
model_all_linear <- ppm(bears_ppp, ~ elev + cover + dist_water + hfi,
                        covariates = list(elev = elev_clean,
                                          cover = cover,
                                          dist_water = dist_water,
                                          hfi = hfi_clean))

print(summary(model_all_linear))
```

```{r}
# Model 2:  Quadratic effects of all 4 covariates
model_all_quad <- ppm(bears_ppp, ~ elev + I(elev^2) + cover + I(cover^2)
                      + dist_water + I(dist_water^2) + hfi + I(hfi^2), 
                      covariates = list(elev = elev_clean,
                                        cover = cover,
                                        dist_water = dist_water,
                                        hfi = hfi_clean))

summary(model_all_quad)
```

```{r}

# Model 3:  Optimal effects of all 4 covariates from the single-covariate analyses
# for forest cover the quadatric model was best
# quadratic for elevation
# dist_water - linear was best
# hfi - quadratic was best

model_all_agg1 <- ppm(bears_ppp, ~ elev + I(elev^2) + cover + I(cover^2)
                      + dist_water + hfi + I(hfi^2), 
                      covariates = list(elev = elev_clean,
                                        cover = cover,
                                        dist_water = dist_water,
                                        hfi = hfi_clean))

summary(model_all_agg1)
```


```{r}

# Model 4:  Aggregate model with following effects
# forest cover - quadatric effect
# elevation - quadratic 
# dist_water - linear
# hfi - linear

model_all_agg2 <- ppm(bears_ppp, ~ elev + I(elev^2) + cover + I(cover^2)
                      + dist_water + hfi, 
                      covariates = list(elev = elev_clean,
                                        cover = cover,
                                        dist_water = dist_water,
                                        hfi = hfi_clean))

summary(model_all_agg2)
```


#### Model evaluation

```{r}
# Compare models using AIC

# Null model
AIC(mod_distw_clean_null)

# Model 1
AIC(model_all_linear)

# Model 2
AIC(model_all_quad)

# Model 3
AIC(model_all_agg1)

# Model 4
AIC(model_all_agg2)

```
**Interpretation:**

- Model 2 (All quadratic effects) has lowest AIC. However, this model produces an error indicating the Fisher information matrix is singular, which typically indicates the existence of multicollinearity or linear dependence among your covariates in the model.
- Therefore we chose next best model (next lowest AIC) which is Model 3 consisting of quadratic effects for Forest cover, Elevation, HFI and linear effects for Distance to Water.

```{r}
# Plot the log intensity using default plot method, by setting log = TRUE
plot(model_all_agg1, se = FALSE, log = TRUE, n = 200,box = F, main = "Log-Intensity Prediction")

#Overlay the bear occurences
plot(bears_ppp,
     pch = 16,
     cex = 0.8,
     cols = "black",
     add = TRUE)
plot(bears_ppp,
     pch = 16,
     cex = 0.5,
     cols = "white",
     add = TRUE)

```

```{r}
#Run the quadrat test
quadrat.test(model_all_agg1, nx = 4, ny = 2)

```
**Interpretation** - The small p value( \< 2.2e-16) indicates that observed data deviate significantly from the model predictions.


```{r}
#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(model_all_agg1, "elev")

#Calculate the relative intensity as a function of forest cover
par_res_forest <- parres(model_all_agg1, "cover")

#Calculate the partial residuals as a function of elevation
par_res_distw <- parres(model_all_agg1, "dist_water")

#Calculate the relative intensity as a function of forest cover
par_res_hfi <- parres(model_all_agg1, "hfi")

#Side by side plotting
par(mfrow = c(2,2))
par(mar = c(4, 4, 2, 2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "Partial residual for elevation",
     xlab = "Elevation (m)")
plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "Partial residual for forest cover",
     xlab = "Forest cover (%)")
plot(par_res_distw,
     legend = FALSE,
     lwd = 2,
     main = "Partial residual for distance to water",
     xlab = "Distance to Water (m)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "Partial residual for HFI",
     xlab = "HFI")
```

### Seasonal Multi-covariate effects

```{r}
options(scipen = 999) # remove scientific notation

# model for each season (individually)
models_seasonal_overall <- list()

for (season in names(ppp_list)) {
  models_seasonal_overall[[season]] <- ppm(ppp_list[[season]], ~ elev + I(elev^2) + 
                                             cover + I(cover^2) + 
                                             dist_water + hfi + I(hfi^2), 
                                        covariates = list(elev = elev_clean,
                                        cover = cover,
                                        dist_water = dist_water,
                                        hfi = hfi_clean))

  cat("Model for", season, ":\n")
  print(models_seasonal_overall[[season]])
}
```

**Interpretation**

-   Winter: Model did not converge, potentially due to collinearity or sparse points. Most variables significant, but I(cover\^2) and dist_water are not.

-   Spring/Summer/Autumn: Models converged well. All have a strong positive effect from hfi, and a curvilinear response to elevation and cover.

#### Model Evaluation

```{r}

for (season in names(models_seasonal_overall)) {

  model <- models_seasonal_overall[[season]]
  #Calculate the partial residuals as a function of elevation
  par_res_elev <- parres(model, "elev")
  
  #Calculate the relative intensity as a function of forest cover
  par_res_forest <- parres(model, "cover")
  
  #Calculate the partial residuals as a function of elevation
  par_res_distw <- parres(model, "dist_water")
  
  #Calculate the relative intensity as a function of forest cover
  par_res_hfi <- parres(model, "hfi")
  
  
  #Side by side plotting
  par(mfrow = c(2,2))
  par(mar = c(4, 4, 2, 2))
  plot(par_res_elev,
       legend = FALSE,
       lwd = 2,
       main = paste("Partial residual for elevation\n",season),
       xlab = "Elevation (m)")
  plot(par_res_forest,
       legend = FALSE,
       lwd = 2,
       main = paste("Partial residual for forest cover\n",season),
       xlab = "Forest cover (%)")
  plot(par_res_distw,
       legend = FALSE,
       lwd = 2,
       main = paste("Partial residual for distance to water\n",season),
       xlab = "Distance to Water (m)")
  plot(par_res_hfi,
       legend = FALSE,
       lwd = 2,
       main = paste("Partial residual for HFI\n", season),
       xlab = "HFI")
 
}
```
